---
title: "Problem Set 3"
author: "Alexander Newton"
date: "12/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(magrittr)
```

## Q5

You have to be kidding me about these data formats!

### a)

```{r}
# as tidy data:

goods_price_quantity <- tibble::tibble(year=c(0,0,1,1),
               good=c("A","B","A","B"),
               pit=c(20,10,10,20),
               qit=c(10,20,20,10),
)

goods_price_quantity %>% dplyr::group_by(year) %>% dplyr::summarise(expenditure = sum(pit*qit))

```

### b)

If instead the quantities for both years were the quantities for year 0:

```{r}
year0_quantities <- goods_price_quantity
year0_quantities[3:4,c("qit")] <- year0_quantities[1:2,c("qit")]
year0_quantities %>% dplyr::group_by(year) %>% dplyr::summarise(expenditure = sum(pit*qit))
```

and as if the quantities were those for year 1:
```{r}
year1_quantities <- goods_price_quantity
year1_quantities[1:2,c("qit")] <- year1_quantities[3:4,c("qit")]
year1_quantities %>% dplyr::group_by(year) %>% dplyr::summarise(expenditure = sum(pit*qit))

```

### c)

calculating inflation for both of the above:

```{r}
df_list <- list("year_0_quantities" = year0_quantities, "year_1_quantities" = year1_quantities)

calculate_inflation <- function(df){
  df %>% dplyr::group_by(year) %>% dplyr::summarise(expenditure = sum(pit*qit)) %>% dplyr::mutate(inflation = expenditure / dplyr::lag(expenditure))
}

lapply(X = df_list, FUN = calculate_inflation)
```

### d)

In the first case inflation was positive, in the second negative. This shows how important it is to have a consistent definition of inflation. The values of inflation were also the inverse of each other: $\dfrac{5}{4}$ against $\dfrac{4}{5}$.

### e)

In the following:

$$
\dfrac{
\sum_i P_{i1} Q_{i1}
}{
\sum_i P_{i0} Q_{i0}
} = 
\dfrac{\sum_i P_{i1} Q_{i0}}{\sum_i P_{i0}Q_{i0}} \times \dfrac{\sum_i P_{i0} Q_{i1}}{\sum_i{P_{i0} Q_{i0}}}
$$

the left term is the expenditure inflation as it is the total expenditure in the second year over the total expenditure in the first. The first term on the right is the price inflation as it is the proportional change in price for a constant quantity, and the second term on the right is the quantity inflation as it is the versa.

The relationship between them is saying that expenditure inflation is the product of quantity inflation and price inflation, in much the same way that expenditure is the product of price and quantity.

## Q6

UK GDP in current market prices in 1995 was £712,548m, while in 1997 it was £801,972m. GDP at constant 1995 market prices in 1997 was £756,144m.

### a) 
Construct index numbers, 1995=100 for: current price GDP; constant price GDP; and the GDP deflator in 1997.

```{r}
GDP <- tibble::tibble(
  year = c(1995, 1997),
  current_price_GDP = c(712548, 801972),
  `1995_price_GDP` = c(712548, 756144)
)

GDP <- dplyr::mutate(GDP,
                     current_price_index_number = current_price_GDP * 100 / GDP[[1,2]],
                     constant_price_index_number = `1995_price_GDP` * 100 / GDP[[1,3]],
                     GDP_deflator = current_price_GDP  / (`1995_price_GDP` ))

GDP
```

### b) 
From these numbers calculate the average annual rate of inflation between 1995 and 1997.

```{r}

total_inflation <- GDP[[ which(GDP$year == 1997), "current_price_index_number"]] / GDP[[ which(GDP$year == 1997), "constant_price_index_number"]] # equal to GDP deflator

average_inflation <- total_inflation ^ 0.5

average_inflation
```

### c)
From these numbers calculate the average annual rate of growth between 1995 and 1997.

```{r}
total_rate_of_growth <- GDP[[ which(GDP$year == 1997), "1995_price_GDP"]] / GDP[[ which(GDP$year == 1995), "1995_price_GDP"]]

average_rate_of_growth <- total_rate_of_growth ^ 0.5

average_rate_of_growth
```


### d)

If the interest rate on two year bonds in 1995 was 10% per annum what would the real per annum interest rate over this period be.

```{r}
current_rate_interest <- 1.1
constant_rate_interest <- current_rate_interest / average_inflation

constant_rate_interest
```

### e)

Explain what Gross Domestic Product measures. What limitations does it have as a measure of the economic wellbeing of a nation.
What other measures are there of inflation or the economic wellbeing of a nation?
How is it measured in different countries?


Gross Domestic Product is a measure of the total output of an economy. It is the total monetary or market value of all the finished goods and services produced within a country's borders in a specific time period. It is calculated by summing the total value added across all firms producing in a country. This is equivalent to the total output minus any intermediate consumption by firms.

GDP movement can be divided into two components: the change in real GDP (which is a GDP in terms of the volume of goods produced), and a GDP deflator which represents an indicator of the change in prices.